---
title: "ABACBS EMCR Hackathon - 1. Obtain annotated scRNA-seq pancreas reference"
author: "Shila Ghazanfar"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 10 
                     fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      cache.lazy = FALSE)

set.seed(2024)
```

# Load packages and data
Load scripts and packages.

```{r}
library(SingleCellExperiment) # for data structure
library(zellkonverter) # for reading scRNAseq h5ad data
library(scran) # for normalisation
library(scater) # for visualisation
library(randomForest) # for predictive model
```

In this document we take an annotated scRNA-seq pancreas dataset from the 
Tabula Sapiens compendium, downloaded from the 
[cellxgene](https://cellxgene.cziscience.com/) website. The direct download
link can be found [here](https://datasets.cellxgene.cziscience.com/9af4585d-a1aa-4109-913f-6fe73772af70.h5ad).
The data object is placed in the relative location `../rawData/`.

We read the scRNA-seq data using the `readH5AD()` function to obtain a 
SingleCellExperiment object, assign rownames as gene symbols, normalise the 
counts according to overall library size, and then subset to features that are 
present in the pancreas MoleculeExperiment object.

```{r}
me_feats = readRDS("../processedData/xenium_pancreas_features.Rds")
sce = readH5AD("../rawData/9af4585d-a1aa-4109-913f-6fe73772af70.h5ad",
               X_name = "counts")
sce

rownames(sce) <- make.unique(as.character(rowData(sce)$feature_name))
sce <- logNormCounts(sce)

feats = intersect(me_feats, rownames(sce))
length(feats)

sce <- sce[feats,]

sce
```

Examine the cell type annotation within this dataset with the reduced number of 
features.

```{r, fig.width = 15, fig.height = 10}
pancreas_colours = setNames(c("#A83279", "#4C9F70", "#E5D352", "#BB77C1", "#42A1F7", 
                              "#F98B60", "#7D59CE", "#FFD23F", "#6D7F99", "#63B7AF",
                              "#EF6F6C", "#46B5B3", "#C1B2AB", "#D6E3C4", "#FF677D"),
                            sort(unique(sce$cell_type)))

sort(table(sce$cell_type))

sce <- runPCA(sce, ncomponents = 15)
sce <- runUMAP(sce)

g = plotReducedDim(sce, "UMAP", colour_by = "cell_type") + 
  scale_colour_manual(values = pancreas_colours)
g
```

# Fit cell type prediction model

We will fit a relatively simple Random Forest model to predict cell type using
the logcounts of the scRNA-seq data. We will use only the features that are 
common between the scRNA-seq and Xenium data.

```{r}
fit = randomForest(x = t(as.matrix(assay(sce, "logcounts"))),
                   y = sce$cell_type)
fit
```

# Save the objects

We save the SingleCellExperiment object in the `../ProcessedData/` folder. We 
also save the fitted Random Forest object and place it in the `../analysisOutput/`
folder.

```{r}
saveRDS(sce, file = "../processedData/sc_pancreas_sce.Rds")
saveRDS(fit, file = "../analysisOutput/sc_cell_type_RF_fit.Rds")
saveRDS(pancreas_colours, file = "../analysisOutput/pancreas_colours.Rds")
```

# Finish

```{r}
sessionInfo()
```
